\bf Interactive Visualization of Hierarchically Structured Data

### 2017-07-12 treelapse.tex
Add description of contributions / context

\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

@@

Our key contribution is the adaptation of the visualization principles of
focus-plus-context and linking to the study of tree-structured data
\citep{buja1996interactive, becker1987brushing}. Our motivating application is
to the analysis of bacterial time series, where an evolutionary tree relating
bacteria is available a priori. However, we have identified common problem types
where, if a tree is not directly available, it can be constructed from data and
then studied using our techniques.

@@

@@

Now that we have specified the essential questions of interest, we survey some
ideas from the visualization literature that can be applied to answer them. As
the core difficulty is high-dimensionality, so it should be no surprise that the
techniques we adapt come from the literature on high-dimensional data
visualization, namely, focus-plus-context and linking.

The focus-plus-context principle is that large collections of visual elements
can be studied at multiple scales, by simultaneously focusing" on a few elements
of interest and maintaining the ``context'' of the coarser-scale background. A
simple example of this idea is to include a search box that highlights matching
samples (focus) and mutes the rest (context). Two more sophisticated methods
anchored in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
central to the proposals in treelapse \citep{hochheiser2004dynamic,
  heer2004doitrees}. In timeboxes, a collection of time series are graphically
queried using interactive brushes. Series that pass through all of the
user-specified brushes are highlighted, and the rest are faded to the
background. Hence, time series meeting the constraints imposed by the brushes
are focused, while the remainder are de-emphasized, though they remain present
as context. This method can be interpreted programmatically as the visual analog
of a database query, or probabilistically as the conditional distribution for
the full series, given it passes through certain bounds.

@@

Further, note that in this subtree averages view, the tree display has changed.
This is because, at each branching point, we place the node with larger average
value on the left. Figure \ref{fig:fig:verrucomicrobiae} notes that the nodes at
the far left in the tree are associated with phylum Verrucomicrobiae,
corresponding to a large average abundance across time points. This phylum had
been previously obfuscated -- because there are not many leaves associated with
this phylum, the sum was small. Interestingly, the abundance of these bacteria
seems to \emph{increase} after the first antibiotics treatment. Be cautious,
however, that the average over only a few Verrucomicrobiae species will be a
more variable estimate than the averages over the more prevalent phyla.

@@

In summary, we have built an easily accessible R package for visualization
techniques in a very specific methodology problem -- analysis of differential
abundance and dynamics in hierarchically structured data -- that appears in a
variety of application domains. We have leveraged a link between R and D3
\citep{vaidyanathan2014htmlwidgets} to create visualizations during the
exploratory phase of data analysis; in this way our work is a departure from the
culture of polished, journalistic visualizations prioritized by the D3
community. Finally, we have given a series of examples to demonstrate how the
general visualization techniques of focus-plus-context and linked brushing can
be practically incorporated into a range of practical analysis workflows, from
studying the impact of bacteria on human health to better allocating units in
commuter bikesharing systems.

>>>>

\usepackage{amsmath, amssymb, amsfonts}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

@@

In particular, we adapt the visualization principles of focus-plus-context and
linking to the study of tree-structured data \citep{buja1996interactive,
  becker1987brushing}. Our motivating application is to the analysis of
bacterial time series, where an evolutionary tree relating bacteria is available
a priori. However, we have identified common problem types where, if a tree is
not directly available, it can be constructed from data and then studied using
our techniques.

@@

In summary, our key contributions are,
\begin{itemize}
\item Proposals for visualizing hierarchically structured data, based on
  principles from the data visualization community.
\item The implementation of these proposals in a publicly available R package.
\item The illustration of the wide reach of hierarchical data visualization,
  through case studies in both scientific and societal contexts.
\end{itemize}

@@

Now that we have specified the essential questions of interest, we survey some
ideas from the visualization literature that can be applied to answer them. As
the core difficulty is high-dimensionality, it should be no surprise that the
techniques we adapt come from the literature on high-dimensional data
visualization, which has enjoyed rapid progress in the last 25 years. Modern
research in this domain develops abstractions and taxonomies for guiding
visualization designs so that they most effectively communicates properties of
the data to their intended audience. A major push in this body of work explores
the potential for interactivity to improve many stages of the data analysis
process, from preliminary data preparation, to refinement and navigation across
views, to final sharing and annotation of results \citep{heer2012taxonomy}.
Further work has attempted to bridge the gap between statistical analysis and
data visualization methodology, both of which provide techniques for learning
from high-dimensional data \citep{de2003visual}.

The problem structures most relevant to our study are tree and temporal
structure, and the visualization community has various ways of reasoning about
these data, see \citep{graham2010survey, aigner2011visualization} for detailed
surveys. From this literature, our approach is most directly informed by the
focus-plus context and linking principles, which we briefly review here. The
focus-plus-context principle is that large collections of visual elements can be
studied at multiple scales, by simultaneously focusing" on a few elements of
interest and maintaining the ``context'' of the coarser-scale background. A
simple example of this idea is to include a search box that highlights matching
samples (focus) and mutes the rest (context). Two more sophisticated methods
anchored in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
central to the proposals in treelapse \citep{hochheiser2004dynamic,
  heer2004doitrees}. In timeboxes, a collection of time series are graphically
queried using interactive brushes. Series that pass through all of the
user-specified brushes are highlighted, and the rest are faded to the
background. Hence, time series meeting the constraints imposed by the brushes
are focused, while the remainder are de-emphasized, though they remain present
as context. This method can be interpreted programmatically as the visual analog
of a database query, or probabilistically as the conditional distribution for
the full series, given it passes through certain bounds.

@@

Further, note that in this subtree averages view, the tree display has changed.
This is because, at each branching point, we place the node with larger average
value on the left. Figure \ref{fig:verrucomicrobiae} notes that the nodes at
the far left in the tree are associated with phylum Verrucomicrobiae,
corresponding to a large average abundance across time points. This phylum had
been previously obfuscated -- because there are not many leaves associated with
this phylum, the sum was small. Interestingly, the abundance of these bacteria
seems to \emph{increase} after the first antibiotics treatment. Be cautious,
however, that the average over only a few Verrucomicrobiae species will be a
more variable estimate than the averages over the more prevalent phyla.

@@

In summary, we have built an easily accessible R package for visualization
techniques in a very specific methodology problem -- analysis of differential
abundance and dynamics in hierarchically structured data -- that appears in a
variety of application domains. We have leveraged a link between R and D3
\citep{vaidyanathan2014htmlwidgets} to create visualizations during the
exploratory phase of data analysis; in this way our work is a departure from the
culture of polished, journalistic visualizations prioritized by the D3 community
and is more closely aligned with the vision in \citep{de2003visual} of more
tightly integrating data visualization and statistical analysis techniques.
Finally, we have given a series of examples to demonstrate how the general
visualization techniques of focus-plus-context and linked brushing can be
practically incorporated into a range of practical analysis workflows, from
studying the impact of bacteria on human health to better allocating units in
commuter bikesharing systems.
